---
title: "Activity Monitoring- Reproducible Research"
author: "Shikha Chamoli"
output: md_document
          
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(xtable)
library(lattice)  
library(lubridate)  

```
##Load & Preprocess data
Read csv file activity.csv in R , check it's summary and preprocess it.

```{r, , echo=TRUE }
daFr <-read.csv("activity.csv")
attach(daFr)
summary(daFr)
daFr$date <- as.Date(daFr$date, "%Y-%m-%d")
head(daFr)

```

##What is mean total number of steps taken per day?

* Calculate the total number of steps taken per day
* Make a histogram of the total number of steps taken each day
* Calculate and report the mean and median of the total number of steps taken per day

```{r , echo=TRUE}
totStepPerDay <- aggregate(list(steps = daFr$steps), list(date = factor(daFr$date)),sum)
head(totStepPerDay)

hist(totStepPerDay$steps,ylim = c(0,30),xlab ="Total steps walked per day" ,
     main= "Total number of steps taken each day! ",  breaks = 10, col="Midnight blue") 

```

```{r , echo=TRUE}
meanStps <- aggregate(list(steps = daFr$steps), list(date = factor(daFr$date)), mean)

medianStps <- aggregate( list(steps = daFr$steps), list(date = factor(daFr$date)), median)

ts <- tapply(daFr$steps, daFr$date, sum)
mean(ts, na.rm = TRUE)
median(ts, na.rm = TRUE)

head(meanStps)
head(medianStps)

plot(meanStps$steps ~ meanStps$date, type= "l", pch=10,xlab="Date" , ylab= "Mean num of steps" , main="Mean num of steps taken on each date!")
  lines(meanStps$steps, type="l", col="midnight blue")

```

##What is the average daily activity pattern?
* Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
* Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r , echo=TRUE}
  
ms<-  tapply(daFr$steps, daFr$interval, mean, na.rm = TRUE)

plot(row.names(ms), ms, type = "l", xlab = "Time Intervals (5-minute)", 
    ylab = "Mean number of steps taken each day!)", main = "Average steps taken at 5 min intervals",     col = "midnight blue")

int_num <- which.max(ms)
intervalWithMaxSteps <- names(int_num)
intervalWithMaxSteps
```

##Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

* Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

```{r , echo=TRUE}

missingVals<- sum(is.na(daFr))
indx_na <- which(is.na(daFr))
imputedVals <- ms[as.character(daFr[indx_na, 3])]
names(imputedVals) <- indx_na
for (i in indx_na) {
    daFr$steps[i] = imputedVals[as.character(i)]
}
sum(is.na(daFr))
par(mfrow=c(1,1))
totStepPerDayAfterImput <- aggregate(list(steps = daFr$steps), list(date = factor(daFr$date)),sum)
head(totStepPerDayAfterImput )
hist(totStepPerDayAfterImput$steps, ylim = c(0,30),xlab ="Total steps walked per day" ,
     main= "Total number of steps taken each day! ",  breaks = 10, col="Midnight blue") 
```

###Conclusion:
* The total number of missing values in the dataset (i.e. the total number of rows with NAs)
      + `r missingVals`  missing values in the column "steps"
* Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
      + I imputed the NA values with the mean of steps. 
* Create a new dataset that is equal to the original dataset but with the missing data filled in.
      + I updated the original dataframe daFr.
* Do these values differ from the estimates from the first part of the assignment? 
      + No
* What is the impact of imputing missing data on the estimates of the total daily number of steps?
      + The mean of overall data distribution has increased in modified dataset.
```{r , echo=TRUE}
daFr$day<- weekdays(as.POSIXct(daFr$date), abbreviate = T)
weekendData<- subset(daFr, daFr[,4] == c("Sun", "Sat"))
weekdayData<- subset(daFr, daFr[,4] != c("Sun", "Sat"))

weekendSteps<- aggregate(steps ~ interval , data=weekendData, FUN=mean)
weekdaySteps<- aggregate(steps ~ interval , data=weekdayData, FUN=mean)

summary(weekendSteps)
summary(weekdaySteps)

par(mfrow=c(2,1))
par(cex = 0.6)

plot(weekdaySteps$steps, type= "l", xlim= c(0,300) ,ylim= c(0,300) ,xlab="Interval" , ylab= "Mean number of steps on weekdays" , main="Number of steps taken per 5-minute interval across weekdays!")
  lines(weekdaySteps$steps, type="l", col="Navy blue")

plot(weekendSteps$steps, type= "l", xlim= c(0,300), ylim= c(0,300), xlab="Interval" , ylab= "Mean number of steps on weekends" , main="Number of steps taken per 5-minute interval across weekends!")
  lines(weekendSteps$steps, type="l", col="Navy blue")

```




###Conclusion:
* We can see from both the graphs above that activity levels on weekdays are much higher and frequent as compared to weekends hence the number of steps is lesser on weekends.  
